---
title: "SIMOA FINAL"
author: "PO Couture"
format: html
editor: visual
---

## SIMOA FINAL

*** Planned Analyses ***
All statistical analyses will be conducted using multiply imputed data to account for missingness across variables. Multiple imputation is used to reduce bias and improve statistical power by generating plausible values based on the observed data distribution. The imputation model includes all variables of interest, consistent with current best practices for handling missing data in psychological and health research.

Group Comparisons
To examine whether older adults who discontinued their use of benzodiazepine receptor agonists (BZRA) differ from those who continued use, a series of independent samples t-tests will be conducted. These analyses will compare the two groups on a range of psychological and personality variables, including:

BFI-10 (Big Five Inventory - 10-item short form),

SURPS (Substance Use Risk Profile Scale),

PHQ-2 (Patient Health Questionnaire – depression screener),

OSSS-3 (Oslo Social Support Scale),

DBAS-16 (Dysfunctional Beliefs and Attitudes about Sleep),

CISS-21 (Coping Inventory for Stressful Situations).

Additional between-group comparisons will be conducted on demographic and health-related variables (e.g., age, gender, sleep quality, comorbidities) to identify any systematic differences that may be associated with BZRA cessation status.

*** Predictive Modelling ***
To identify the most important variables associated with successful BZRA cessation, a Random Forest classification model will be employed. Random Forest is a non-parametric ensemble machine learning method that handles complex interactions and non-linear relationships, and is robust to multicollinearity and overfitting.

The Random Forest model will be trained using all personality, psychological, demographic, and health-related variables as predictors, with BZRA cessation (yes/no) as the outcome. Variable importance scores will be used to rank predictors based on their contribution to classification accuracy.

*** Model Confirmation ***
To validate the findings from the Random Forest model, a logistic regression analysis will be conducted using the top predictors identified by the Random Forest. This traditional regression model will allow for the estimation of effect sizes (odds ratios) and the statistical significance of each variable’s unique contribution to BZRA cessation. Confidence intervals and p-values will be reported, and model diagnostics will be used to assess model fit.

Together, this multi-step analytic strategy aims to both explore and confirm key predictors of BZRA discontinuation among older adults, leveraging the strengths of both machine learning and traditional inferential statistics.

## Data Loading, Screening and Packages
In this section I am loading all the necessary packages for my analysis and loading in the data.
```{r}
#| label: Data Loading and Packages

# Installing and Loading Packages
install.packages("MissMech")

library(dplyr)
library(ggplot2)
library(mice)
library(MissMech)
library(naniar)
library(readr)
library(tidyverse)
library(VIM)

# Data Loading
SIMOA_Report <- read_csv("SIMOA Report.csv")
View(SIMOA_Report)

```

Below I am creating a new object containing only the participants we can confirm are >= 65 and are using a BZRA by answering which specific BZRA they are using. This is to ensure they are not using other medications that may have sedative effects such as antihistamines or SSRI's.
```{r}
#| label: Creating new object 

Dataset <- SIMOA_Report %>%
  # Apply filtering based on age_cat and age
  filter(
    age_cat == 1 | (age_cat == 0 & age >= 65)
  ) %>%
  # Apply filtering to keep only rows where any of c_sp___1 to c_sp___14 == 1
  filter(
    rowSums(select(., starts_with("c_sp___"))[, 1:14] == 1, na.rm = TRUE) > 0
  )
```

## Data Cleaning
Below is the section where I will get rid of all the columns where nobody answered the question and all the values are NA. Alternatively I will try to divide based on who stopped and who did not and then from there remove the columns which they did not have the opportunity to answer.
```{r}
#| label: Data Cleaning

#This section will group the questions into which ones were seen by those who answered yes to the "scrn_stopped_bzra" question and have stopped BZRA use.
yes_questions <- c""
```


## Multiple Imputation 
```{r}
#| label: 


```








